---
title: "Metagenomics_IL10_ZnS_ZnA_ZnS"
output: html_document
date: "2025-02-21"
---

#LOAD METADATA

```{r}

metadata_IL10 <- read.table("/Users/savitasastry/Downloads/Zinc_Deficiency_SS/Metagenomics/IL10_ZnD_ZnA_ZnS_metadata.csv", header = TRUE, sep = ",", stringsAsFactors = FALSE)
metadata_IL10$Treatment <- as.factor(metadata_IL10$Treatment)
metadata_IL10$Treatment <- relevel(metadata_IL10$Treatment, ref = "ZnA")
metadata_IL10
```

#COMBINED DATA FOR BRACKEN USING SPECIES NAMES AS THE ROW NAMES

```{r}

library(data.table)
library(dplyr)
# List of Bracken files
file_list <- list.files(path = "/Users/savitasastry/Downloads/Zinc_Deficiency_SS/Metagenomics/braken_outputs/IL10:IL10+Zn/Species_Level/", pattern = "_bracken_output.txt", full.names = TRUE)

# Read each file and process
bracken_IL10 <- lapply(file_list, function(file) {
  # Extract sample name from the file name
  sample_name <- gsub("_bracken_output.txt", "", basename(file))
  
  # Read the file and keep necessary columns
  data <- read.table(file, header = TRUE, sep = "\t") %>%
    select(name, new_est_reads) %>%
    rename_with(~ sample_name, .cols = new_est_reads)  # Correct way to rename dynamically

  return(data)
})

# Merge all data by the "Taxon" column
combined_bracken_IL10 <- Reduce(function(x, y) merge(x, y, by = "name", all = TRUE), bracken_IL10)

# Replace NA with 0 (if there are missing taxa in some samples)
combined_bracken_IL10[is.na(combined_bracken_IL10)] <- 0

rownames(combined_bracken_IL10) <- combined_bracken_IL10$name
combined_bracken_IL10 <- combined_bracken_IL10[, -1]

write.csv(combined_bracken_IL10, "/Users/savitasastry/Downloads/Zinc_Deficiency_SS/Metagenomics/IL10_Proj/combined_data_IL10.csv", row.names = TRUE)
combined_bracken_IL10 <- read.csv("/Users/savitasastry/Downloads/Zinc_Deficiency_SS/Metagenomics/IL10_Proj/combined_data_IL10.csv", row.names = 1)
colnames(combined_bracken_IL10) <- gsub("^X", "", colnames(combined_bracken_IL10))
combined_bracken_IL10
```


#GENERATE A READS BOXPLOT TO IDENTIFY ANY MAJOR CHANGES IN READ NUMBERS BETWEEN GROUPS. BASED ON THE RESULTS BELOW, THERE ARE NO SIGNIFICANT DIFFERENCES BETWEEN OUR TOTAL READS IN THE DIFFERENT TREATMENT GROUPS. THEREFORE, I WILL NOT BE RAREFYING MY READS

```{r}
library(ggplot2)
library(dplyr)
library(ggpubr)

bracken_long <- combined_bracken_IL10 %>%
  as.data.frame() %>%
  rownames_to_column(var = "Species") %>%
  pivot_longer(cols = -Species, names_to = "Sample", values_to = "Reads")

# Merge with metadata to get Treatment and Location
bracken_long <- bracken_long %>%
  left_join(metadata_IL10, by = c("Sample" = "Sample"))

# Create a combined column for grouping (Treatment + Location)
bracken_long <- bracken_long %>%
  mutate(Group = paste(Treatment, Location, sep = " "))

# 🔹 Summarize total reads per sample
total_reads_per_sample <- bracken_long %>%
  group_by(Sample, Treatment) %>%
  summarise(Total_Reads = sum(Reads, na.rm = TRUE), .groups = "drop")

# 🔹 Create the boxplot with pairwise comparisons
read_dist_IL10 <- ggplot(total_reads_per_sample, aes(x = paste(Treatment, sep = " "), 
                                   y = Total_Reads, fill = Treatment)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6) +  # Boxplot without outliers
  geom_jitter(width = 0.2, size = 2, alpha = 0.8) + # Only 1 dot per sample
  scale_y_log10() +  # Log scale for better visualization
  stat_compare_means(method = "wilcox.test", label = "p.signif", 
                     comparisons = list(
                       c("Chow", "ZnS"),
                       c("ZnA", "ZnD"),
                       c("ZnS", "ZnD")
                     )) +  # Pairwise comparisons
  labs(
    title = "Total Read Counts per Sample by Treatment and Location",
    x = "Group",
    y = "Total Reads (log10 scale)",
    fill = "Treatment"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 12))

read_dist_IL10
ggsave("/Users/savitasastry/Downloads/Zinc_Deficiency_SS/Metagenomics/IL10_Proj/figures/read_dist_IL10.png", width = 10, height = 6, dpi = 300)


```


#MAKE PCA PLOT GRAPHING BRAY CURTIS DISSIMILARITY FOR DIMENSIONALITY REDUCTION

```{r}
# Load required libraries
library(ggplot2)
library(vegan)  # For Bray-Curtis distance
library(dplyr)  # Data manipulation
library(janitor) # Cleaning column names
library(ggrepel) # For sample labels

# 🔹 Merge Bracken data with metadata
pcoa_input_IL10 <- merge(t(combined_bracken_IL10), metadata_IL10, by.x = "row.names", by.y = "Sample")
colnames(pcoa_input_IL10)[1] <- "Sample"  # Ensure first column is Sample
pcoa_input_IL10
# 🔹 Select only numerical abundance data
pcoa_numeric_IL10 <- pcoa_input_IL10[, !(colnames(pcoa_input_IL10) %in% c("Sample", "Treatment"))]

# Convert the matrix to character (in case some columns are factors)
pcoa_numeric_IL10 <- apply(pcoa_numeric_IL10, 2, as.character)

# Trim leading/trailing spaces and convert to numeric
pcoa_numeric_IL10 <- apply(pcoa_numeric_IL10, 2, function(x) as.numeric(trimws(x)))

# Ensure it is a numeric matrix
pcoa_numeric_IL10 <- as.matrix(pcoa_numeric_IL10)
pcoa_numeric_IL10
# Check that the conversion worked
str(pcoa_numeric_IL10)  # Should display 'num' type

# Replace NA values with 0 to avoid division errors
pcoa_numeric_IL10[is.na(pcoa_numeric_IL10)] <- 0

# Compute row sums
row_sums <- rowSums(pcoa_numeric_IL10)

# Check if there are any zero row sums
print(sum(row_sums == 0))  # If this prints >0, you have zero-sum rows!


# 🔹 Normalize by total reads per sample (relative abundance)
pcoa_numeric_IL10 <- pcoa_numeric_IL10 / rowSums(pcoa_numeric_IL10)
pcoa_numeric_IL10
# 🔹 Compute Bray-Curtis distance matrix
bray_curtis_dist <- vegdist(pcoa_numeric_IL10, method = "bray")

# 🔹 Perform PCoA (Principal Coordinates Analysis)
pcoa_result_IL10 <- cmdscale(bray_curtis_dist, k = 2, eig = TRUE)  # Compute eigenvalues

# 🔹 Convert PCoA results into a data frame
pcoa_scores_IL10 <- as.data.frame(pcoa_result_IL10$points)
colnames(pcoa_scores_IL10) <- c("PC1", "PC2")
pcoa_scores_IL10$Sample <- pcoa_input_IL10$Sample

# 🔹 Merge PCoA scores with metadata
pcoa_scores_IL10 <- merge(pcoa_scores_IL10, metadata_IL10, by = "Sample")

# 🔹 Calculate variance explained
pc1_variance_IL10 <- round(100 * (pcoa_result_IL10$eig[1] / sum(pcoa_result_IL10$eig)), 1)
pc2_variance_IL10 <- round(100 * (pcoa_result_IL10$eig[2] / sum(pcoa_result_IL10$eig)), 1)

# 🔹 Print variance explained
cat("PC1 Variance:", pc1_variance_IL10, "%\n")
cat("PC2 Variance:", pc2_variance_IL10, "%\n")

# 🔹 Create a new Grouping variable: "ZnD/ZnA" + "Location"
pcoa_scores_IL10$Group <- paste(pcoa_scores_IL10$Treatment, sep = " ")

# 🔹 Define colors for the 6 groups
group_colors <- c(
  "ZnD" = "#E69F00", "ZnA" = "#56B4E9", "ZnS" = "#009E73", "Chow" = "#F0E442")

pcoa_scores_IL10
# 🔹 PCoA Plot: ZnD/ZnA Groups
pcoa_plot_IL10 <- ggplot(pcoa_scores_IL10, aes(x = PC1, y = PC2, color = Group, fill = Group)) +
  geom_point(size = 3, alpha = 0.8) +
  stat_ellipse(geom = "polygon", alpha = 0.2) +  # Add filled ellipses
  scale_color_manual(values = group_colors) +
  scale_fill_manual(values = group_colors) +
  theme_minimal() +
  labs(
    title = "PCoA (Bray-Curtis) - IL10 Groups",
    x = paste0("PC1 (", pc1_variance_IL10, "%)"),
    y = paste0("PC2 (", pc2_variance_IL10, "%)"),
    color = "Group"
  ) +
  theme(
    legend.position = "right",
    panel.background = element_rect(fill = "white", color = "black"),
    plot.title = element_text(face = "bold", size = 14),
    axis.title.x = element_text(face = "bold", size = 12),
    axis.title.y = element_text(face = "bold", size = 12)
  )

# 🔹 Save and print the plot
ggsave("/Users/savitasastry/Downloads/Zinc_Deficiency_SS/Metagenomics/IL10_Proj/figures/pcoa_IL10.png",
       plot = pcoa_plot_IL10, width = 8, height = 6, dpi = 300)
pcoa_plot_IL10

```

#DATAFRAME FROM KRAKEN OUTPUT TO CALCULATE THE ABUNDANCE OF BACTERIA, FUNGI, AND VIRUSES ACROSS SAMPLES

```{r}
library(dplyr)

# Set the folder path
input_folder <- "/Users/savitasastry/Downloads/Zinc_Deficiency_SS/Metagenomics/kraken_outputs/IL10:IL10+Zn/"

# List all Kraken report files in the folder
file_list <- list.files(path = input_folder, pattern = "*_report.txt", full.names = TRUE)

# Initialize an empty list to store results
taxon_counts <- list()

# Loop through each Kraken report file
for (file in file_list) {
  
  # Read the Kraken report file
  kraken_report <- read.delim(file, header = FALSE, sep = "\t", stringsAsFactors = FALSE)
  
  # Rename columns (adjust as needed based on the structure of your file)
  colnames(kraken_report) <- c("Percent_Reads", "Reads_This_Taxon", "Reads_Clade", "Rank_Code", "Taxonomy_ID", "Taxon_Name")
  
  # Filter for taxa of interest
  filtered_taxa <- kraken_report %>%
    filter(Taxon_Name %in% c("    Bacteria", "  Viruses", "        Fungi"))
  
  # Aggregate counts for each taxon
  taxon_summary <- filtered_taxa %>%
    group_by(Taxon_Name) %>%
    summarize(Read_Count = sum(Reads_This_Taxon), .groups = "drop")
  
  # Add the sample name (derived from the filename)
  taxon_summary$Sample <- gsub("_report.txt", "", basename(file))
  
  # Append to the list
  taxon_counts[[file]] <- taxon_summary
}

# Combine all results into a single dataframe
combined_taxon_counts <- bind_rows(taxon_counts)

# Normalize counts by total reads per sample
# Calculate total reads per sample
total_reads <- combined_taxon_counts %>%
  group_by(Sample) %>%
  summarize(Total_Reads = sum(Read_Count))

# Join total reads with the combined taxon counts
normalized_taxon_counts <- combined_taxon_counts %>%
  left_join(total_reads, by = "Sample") %>%
  mutate(Relative_Abundance = Read_Count / Total_Reads * 100)

transformed_df_PF <- normalized_taxon_counts %>%
  select(Taxon_Name, Sample, Read_Count) %>% # Keep relevant columns
  pivot_wider(
    names_from = Sample,           # Columns become Sample names
    values_from = Read_Count, # Values are from Relative_Abundance
    values_fill = 0                # Fill missing values with 0
  )

transformed_df_PF <- as.data.frame(transformed_df_PF)
rownames(transformed_df_PF) <- transformed_df_PF$Taxon_Name
transformed_df_PF <- transformed_df_PF[, -1]
# Rename columns: replace "-" with "." and remove "_kraken"
colnames(transformed_df_PF) <- gsub("-", ".", colnames(transformed_df_PF))  # Replace "-" with "."
colnames(transformed_df_PF) <- gsub("_kraken$", "", colnames(transformed_df_PF))  # Remove "_kraken" at the end
transformed_df_PF <- transformed_df_PF[, colnames(transformed_df_PF) %in% metadata_IL10$Sample]


write.csv(transformed_df_PF, "/Users/savitasastry/Downloads/Zinc_Deficiency_SS/Metagenomics/IL10_Proj/kingdom_classification_IL10.csv", row.names = TRUE)

kingdom_classification_IL10 <- read.csv("/Users/savitasastry/Downloads/Zinc_Deficiency_SS/Metagenomics/IL10_Proj/kingdom_classification_IL10.csv", row.names = 1)
kingdom_classification_IL10

colnames(kingdom_classification_IL10) <- gsub("^X", "", colnames(kingdom_classification_IL10))
rownames(kingdom_classification_IL10)

```

#PLOT SIMPLE RELATIVE ABUNDANCE GRAPH VISUALIZING THE PRESENCE OF BACTERIAL, FUNGAL, AND VIRAL READS
```{r}
library(ggplot2)
library(dplyr)
library(tidyr)
library(ggbreak)  # Import the package for axis breaks

# Convert read counts to relative abundance (%)
relative_abundance_df <- kingdom_classification_IL10 %>%
  mutate(across(everything(), ~ . / sum(.) * 100)) %>%
  tibble::rownames_to_column(var = "Kingdom") %>%
  pivot_longer(cols = -Kingdom, names_to = "Sample", values_to = "Relative_Abundance")

# Merge with metadata to get treatment groups
relative_abundance_df <- relative_abundance_df %>%
  left_join(metadata_ZnD5, by = c("Sample" = "Sample"))

relative_abundance_df

# Create boxplot with a y-axis break
boxplot_relative_abundance <- ggplot(relative_abundance_df, aes(x = Kingdom, y = Relative_Abundance, fill = Kingdom)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.5) +  # Adds individual points for better visualization
  scale_fill_manual(values = c("    Bacteria" = "#1f78b4", "        Fungi" = "#6a3d9a", "  Viruses" = "#33a02c")) +
  theme_minimal() +
  labs(title = "Relative Abundance of Bacteria, Fungi, and Viruses",
       x = "Kingdom",
       y = "Relative Abundance (%)") +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title.x = element_text(size = 14, face = "bold"),
    axis.title.y = element_text(size = 14, face = "bold"),
    axis.text.x = element_text(size = 12, face = "bold"),
    axis.text.y = element_text(size = 12, face = "bold"),
    legend.position = "none"
  ) +
  scale_y_break(c(3, 95))  # Breaks y-axis between 5 and 90

# Save plot
ggsave("/Users/savitasastry/Downloads/Zinc_Deficiency_SS/Metagenomics/IL10_Proj/figures/relative_abundance_boxplot_with_break.jpeg",
       plot = boxplot_relative_abundance, width = 8, height = 6)

# Display plot
print(boxplot_relative_abundance)

unique(relative_abundance_df$Kingdom)

```


#CALCULATE SHANNON DIVERSITY, CHAO1 DIVERSITY, AND OBSERVED SPECIES IN PLUSPF DATABASE
#Generate phyloseq object and make plots for diversity 

```{r fig.height= 8}
library(phyloseq)
library(dplyr)
library(tidyr)

metadata_IL10_phy <- metadata_IL10
rownames(metadata_IL10_phy) <- metadata_IL10_phy$Sample
metadata_IL10_phy <- metadata_IL10_phy[, -which(colnames(metadata_IL10_phy) == "Sample")]
head(metadata_IL10_phy)
metadata_IL10_phy <- sample_data(metadata_IL10_phy)

otu_phy_IL10 <- combined_bracken_IL10
otu_phy_IL10 <- as.matrix(otu_phy_IL10)
otu_phy_IL10 <- otu_table(otu_phy_IL10, taxa_are_rows = TRUE)

physeq_IL10 <- phyloseq(otu_phy_IL10, metadata_IL10_phy)
diversity_data_IL10 <- estimate_richness(physeq_IL10, measures = c("Observed", "Shannon", "Chao1"))
metadata_IL10 <- sample_data(physeq_IL10)  # Extract sample metadata
diversity_data_IL10 <- cbind(metadata_IL10, diversity_data_IL10)  

```

```{r}
library(phyloseq)
library(dplyr)
library(ggplot2)
library(ggpubr)

# Compute Diversity Metrics
diversity_data_IL10 <- estimate_richness(physeq_IL10, measures = c("Observed", "Shannon", "Chao1"))

# Extract Sample Metadata
metadata_IL10_df <- as.data.frame(sample_data(physeq_IL10))

# Merge Diversity Data with Metadata
diversity_data_IL10 <- cbind(metadata_IL10_df, diversity_data_IL10)

custom_colors <- c(
  "ZnD" = "#E69F00",
  "ZnA" = "#56B4E9",
   "ZnS" = "#009E73",
  "Chow" = "#F0E442"
)

plot_diversity_metric <- function(metric, y_label, title, subset_location) {
  
  # Subset the data for the specific location
  subset_data <- diversity_data_IL10
  subset_data$Treatment <- factor(subset_data$Treatment, levels = c("Chow", "ZnS", "ZnA", "ZnD"))
  
  ggplot(subset_data, aes(x = Treatment, y = get(metric), fill = Treatment)) +
    geom_boxplot(alpha = 0.7, outlier.shape = NA) +
    geom_jitter(width = 0.2, alpha = 0.6, size = 1.5) +  
    scale_fill_manual(values = custom_colors) +
    theme_minimal() +
    labs(title = paste(title, "-"), x = "Treatment Group", y = y_label) +
    theme(
      plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
      axis.title.x = element_text(size = 16, face = "bold"),
      axis.title.y = element_text(size = 16, face = "bold"),
      axis.text.x = element_text(size = 14, face = "bold", color = "black"),
      axis.text.y = element_text(size = 14, color = "black")
    ) +
    stat_compare_means(method = "kruskal.test", label.y = max(subset_data[[metric]], na.rm = TRUE) * 1.1) +  # Log scale for better visualization
  stat_compare_means(method = "wilcox.test", label = "p.signif", 
                     comparisons = list(
                       c("Chow", "ZnS"),
                       c("ZnA", "ZnD"),
                       c("ZnS", "ZnD")
                     ))   # Pairwise comparisons
}

# Shannon Diversity
shannon_IL10 <- plot_diversity_metric("Shannon", "Shannon Diversity Index", "Shannon Diversity")

# Chao1 Index
chao1_IL10 <- plot_diversity_metric("Chao1", "Chao1 Index", "Chao1 Index", "Cecum")

# Observed Species
observed_IL10 <- plot_diversity_metric("Observed", "Number of Observed Species", "Observed Species", "cecum")

# Define output directory
output_dir <- "/Users/savitasastry/Downloads/Zinc_Deficiency_SS/Metagenomics/IL10_Proj/figures/DiversityPlots/"

# Save all plots
ggsave(paste0(output_dir, "shannon_cecum.png"), plot = shannon_IL10, width = 8, height = 6, dpi = 300)
ggsave(paste0(output_dir, "chao1_cecum.png"), plot = chao1_IL10, width = 8, height = 6, dpi = 300)
ggsave(paste0(output_dir, "observed_cecum.png"), plot = observed_IL10, width = 8, height = 6, dpi = 300)

print(shannon_IL10)
print(chao1_IL10)
print(observed_IL10)

```


#GENERATE KRAKEN REPORT FILES FOR BACTERIA, FUNGI, AND VIRUSES SEPARATED. THEN I WILL USE THE SPECIES INFORMATION FROM THE PARSED KRAKEN FILES TO SUBSET THE BRACKEN DATA. THEN I WILL CLUSTER THEM IN A PCA PLOT AND MAKE GRAPHS AND DO DIFFERENTIAL EXPRESSION ANALYSIS ON THEM SEPARATELY.


```{r}
# Load necessary library
library(fs)  # For file handling

# Define input and output directories
input_dir <- "/Users/savitasastry/Downloads/Zinc_Deficiency_SS/Metagenomics/kraken_outputs/IL10:IL10+Zn/"
output_dir <- "/Users/savitasastry/Downloads/Zinc_Deficiency_SS/Metagenomics/IL10_Proj/kraken_output_kingdom_separated/"

# Create output directory if it doesn't exist
if (!dir.exists(output_dir)) {
  dir.create(output_dir, recursive = TRUE)
}

# Function to extract full sections for bacteria, fungi, and viruses
extract_section <- function(data, keyword) {
  start_index <- grep(keyword, data)
  if (length(start_index) == 0) return(character(0))  # Return empty if keyword not found
  
  extracted_lines <- c()
  collecting <- FALSE
  
  for (line in data) {
    if (grepl(keyword, line)) collecting <- TRUE  # Start collecting when we find the keyword
    if (collecting) extracted_lines <- c(extracted_lines, line)
    
    # Stop collecting when a new domain appears (but not in the same category)
    if (collecting && grepl("D\\t", line) && !grepl(keyword, line)) break
  }
  
  return(extracted_lines)
}

# Get all Kraken report files (only those that match the expected report pattern)
kraken_files <- list.files(input_dir, pattern = ".*_report.txt$", full.names = TRUE)

# Loop through each file and process it
for (file in kraken_files) {
  # Read the Kraken report file
  kraken_data <- readLines(file)
  
  # Extract full sections for each domain
  bacteria_lines <- extract_section(kraken_data, "D\\t2\\t")  # Bacteria
  fungi_lines <- extract_section(kraken_data, "K\\t4751\\t")  # Fungi
  virus_lines <- extract_section(kraken_data, "D\\t10239\\t")  # Viruses

  # Generate output file names
  file_name <- basename(file)
  output_bacteria <- file.path(output_dir, paste0(tools::file_path_sans_ext(file_name), "_bacteria.txt"))
  output_fungi <- file.path(output_dir, paste0(tools::file_path_sans_ext(file_name), "_fungi.txt"))
  output_virus <- file.path(output_dir, paste0(tools::file_path_sans_ext(file_name), "_virus.txt"))

  # Write results to files only if data exists
  if (length(bacteria_lines) > 0) writeLines(bacteria_lines, output_bacteria)
  if (length(fungi_lines) > 0) writeLines(fungi_lines, output_fungi)
  if (length(virus_lines) > 0) writeLines(virus_lines, output_virus)

  # Print confirmation message
  cat("Processed:", file_name, "\n")
}

cat("All Kraken reports have been processed. Separated files are saved in:\n", output_dir, "\n")
hi
```


```{r}
# Load necessary libraries
library(dplyr)
library(readr)
library(tools)

# Define directories
bracken_dir <- "/Users/savitasastry/Downloads/Zinc_Deficiency_SS/Metagenomics/braken_outputs/IL10:IL10+Zn/Species_Level"
kraken_sep_dir <- "/Users/savitasastry/Downloads/Zinc_Deficiency_SS/Metagenomics/IL10_Proj/kraken_output_kingdom_separated"
output_dir <- "/Users/savitasastry/Downloads/Zinc_Deficiency_SS/Metagenomics/IL10_Proj/bracken_output_separated"

# Create output directory if it doesn't exist
if (!dir.exists(output_dir)) {
  dir.create(output_dir, recursive = TRUE)
}

# Function to extract taxonomic IDs from a Kraken-separated report
extract_tax_ids <- function(kraken_file) {
  if (!file.exists(kraken_file)) return(character(0))  # Return empty if file does not exist
  kraken_data <- read_tsv(kraken_file, col_names = FALSE, show_col_types = FALSE)
  return(unique(kraken_data$X5))  # Extract taxonomic ID column (assumed in column 5)
}

# Function to subset Bracken output by kingdom
subset_bracken <- function(bracken_data, tax_ids, output_name) {
  filtered_data <- bracken_data %>% filter(taxonomy_id %in% tax_ids)
  write_tsv(filtered_data, output_name)
}

# Get all Bracken files
bracken_files <- list.files(bracken_dir, pattern = "_bracken_output.txt$", full.names = TRUE)

# Process each Bracken file
for (bracken_file in bracken_files) {
  # Extract the base name by removing "_bracken_output" from the filename
  base_name <- gsub("_bracken_output", "", tools::file_path_sans_ext(basename(bracken_file)))
  # Extract base name without extension

  # Define corresponding Kraken-separated files
  kraken_bacteria_file <- file.path(kraken_sep_dir, paste0(base_name, "_kraken_report_bacteria.txt"))
  kraken_fungi_file <- file.path(kraken_sep_dir, paste0(base_name, "_kraken_report_fungi.txt"))
  kraken_virus_file <- file.path(kraken_sep_dir, paste0(base_name, "_kraken_report_virus.txt"))


  # Extract taxonomic IDs for each kingdom
  bacteria_ids <- extract_tax_ids(kraken_bacteria_file)
  fungi_ids <- extract_tax_ids(kraken_fungi_file)
  virus_ids <- extract_tax_ids(kraken_virus_file)

  # Read Bracken output file
  bracken_data <- read_tsv(bracken_file, show_col_types = FALSE)

  # Generate output file paths
  bacteria_output <- file.path(output_dir, paste0(base_name, "_bacteria.tsv"))
  fungi_output <- file.path(output_dir, paste0(base_name, "_fungi.tsv"))
  virus_output <- file.path(output_dir, paste0(base_name, "_virus.tsv"))

  # Filter and save Bracken output for each kingdom
  if (length(bacteria_ids) > 0) subset_bracken(bracken_data, bacteria_ids, bacteria_output)
  if (length(fungi_ids) > 0) subset_bracken(bracken_data, fungi_ids, fungi_output)
  if (length(virus_ids) > 0) subset_bracken(bracken_data, virus_ids, virus_output)

  # Print status message
  cat("Processed:", bracken_file, "\n")
}

cat("All Bracken outputs separated by kingdom saved in:", output_dir, "\n")
```

```{r}
# Load necessary libraries
library(data.table)
library(dplyr)

# Define directories
bracken_sep_dir <- "/Users/savitasastry/Downloads/Zinc_Deficiency_SS/Metagenomics/IL10_Proj/bracken_output_separated/"
output_file_bacteria <- "/Users/savitasastry/Downloads/Zinc_Deficiency_SS/Metagenomics/IL10_Proj/combined_bacteria.csv"
output_file_fungi <- "/Users/savitasastry/Downloads/Zinc_Deficiency_SS/Metagenomics/IL10_Proj/combined_fungi.csv"
output_file_virus <- "/Users/savitasastry/Downloads/Zinc_Deficiency_SS/Metagenomics/IL10_Proj/combined_virus.csv"

# Function to process and merge Bracken files by kingdom
combine_bracken_files <- function(pattern) {
  # List all files matching the pattern (e.g., "_bacteria.tsv")
  file_list <- list.files(path = bracken_sep_dir, pattern = pattern, full.names = TRUE)

  # Read each file and extract relevant columns
  bracken_data <- lapply(file_list, function(file) {
    sample_name <- gsub(pattern, "", basename(file))  # Extract sample name
    data <- read.table(file, header = TRUE, sep = "\t") %>%
      select(name, new_est_reads) %>%
      rename_with(~ sample_name, .cols = new_est_reads)  # Rename column with sample name
    return(data)
  })

  # Merge all data by the "Taxon" column (name)
  combined_data <- Reduce(function(x, y) merge(x, y, by = "name", all = TRUE), bracken_data)

  # Replace NA with 0 (if there are missing taxa in some samples)
  combined_data[is.na(combined_data)] <- 0

  # Set row names as taxa names and remove the "name" column
  rownames(combined_data) <- combined_data$name
  combined_data <- combined_data[, -1]

  return(combined_data)
}

# Process and merge files for each kingdom
combined_bacteria <- combine_bracken_files("_bacteria.tsv")
combined_fungi <- combine_bracken_files("_fungi.tsv")
combined_virus <- combine_bracken_files("_virus.tsv")

# Save as CSV
write.csv(combined_bacteria, output_file_bacteria, row.names = TRUE)
write.csv(combined_fungi, output_file_fungi, row.names = TRUE)
write.csv(combined_virus, output_file_virus, row.names = TRUE)

# Read files back to clean column names
combined_bacteria <- read.csv(output_file_bacteria, row.names = 1)
combined_fungi <- read.csv(output_file_fungi, row.names = 1)
combined_virus <- read.csv(output_file_virus, row.names = 1)

# Remove extra "X" added to column names
colnames(combined_bacteria) <- gsub("^X", "", colnames(combined_bacteria))
colnames(combined_fungi) <- gsub("^X", "", colnames(combined_fungi))
colnames(combined_virus) <- gsub("^X", "", colnames(combined_virus))

# Print confirmation
cat("Combined Bracken data saved:\n",
    output_file_bacteria, "\n",
    output_file_fungi, "\n",
    output_file_virus, "\n")

# Return dataframes
list(
  bacteria = combined_bacteria,
  fungi = combined_fungi,
  virus = combined_virus
)

```


#GENERATE DIVERSITY PLOTS FROM THE STRATIFIED OUTPUT FOR BACTERIA, FUNGI, AND VIRUSES

```{r}


# Load required libraries
library(ggplot2)
library(vegan)
library(dplyr)
library(janitor)
library(ggrepel)
library(phyloseq)
library(tidyr)
library(ggpubr)

# Define directories
data_dir <- "/Users/savitasastry/Downloads/Zinc_Deficiency_SS/Metagenomics/IL10_Proj"
figures_dir <- file.path(data_dir, "figures")
pcoa_fig_dir <- file.path(figures_dir, "PCA")
diversity_fig_dir <- file.path(figures_dir, "DiversityPlots")

# Load metadata
metadata <- metadata_IL10
rownames(metadata) <- metadata$Sample 


compute_diversity <- function(data, kingdom) {
  # **Check and Fix Sample Names**
  sample_names_data <- colnames(data)  # Extract sample names from Bracken data
  sample_names_metadata <- metadata$Sample  # Extract sample names from metadata

  # Find mismatches
  mismatched_samples <- setdiff(sample_names_data, sample_names_metadata)
  if (length(mismatched_samples) > 0) {
    stop("Error: Mismatched sample names found. Check and rename metadata sample names.")
  }

  # Convert data into a numeric matrix
  data_matrix <- as.matrix(data)  # Keep taxa as rows, samples as columns
  mode(data_matrix) <- "numeric"  # Ensure numeric conversion

  # Ensure metadata row names match Bracken sample names
  rownames(metadata) <- metadata$Sample
  metadata_phy <- sample_data(metadata)  # Convert metadata to phyloseq object

  # Convert OTU table into a `phyloseq` format
  otu_table_phy <- otu_table(data_matrix, taxa_are_rows = TRUE)
  physeq <- phyloseq(otu_table_phy, metadata_phy)

  # Check if sample names match
  if (!all(sample_names(metadata_phy) %in% sample_names(otu_table_phy))) {
    stop("Error: Sample names in metadata do not match sample names in OTU table.")
  }

  # Compute diversity metrics
  diversity_data <- estimate_richness(physeq, measures = c("Observed", "Shannon", "Chao1"))
  diversity_data <- cbind(metadata, diversity_data)
  

  # **Define function to plot diversity metrics**
  plot_diversity_metric <- function(metric, y_label, title) {
    ggplot(diversity_data, aes(x = Treatment, y = get(metric), fill = Treatment)) +
      geom_boxplot(alpha = 0.7, outlier.shape = NA) +
      scale_fill_manual(values = custom_colors) +
      geom_jitter(width = 0.2, alpha = 0.6, size = 1.5) +
      theme_minimal() +
      labs(title = title, x = "Treatment Group", y = y_label) +
      theme(
        plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
        axis.title.x = element_text(size = 16, face = "bold"),
        axis.title.y = element_text(size = 16, face = "bold"),
        axis.text.x = element_text(size = 14, face = "bold", color = "black"),
        axis.text.y = element_text(size = 14, color = "black")
      ) +
      stat_compare_means(method = "kruskal.test")  # Kruskal-Wallis test
  }

  # Generate plots
  shannon_plot <- plot_diversity_metric("Shannon", "Shannon Diversity Index", paste("Shannon Diversity -", kingdom))
  chao1_plot <- plot_diversity_metric("Chao1", "Chao1 Index", paste("Chao1 Index -", kingdom))
  observed_plot <- plot_diversity_metric("Observed", "Number of Observed Species", paste("Observed Species -", kingdom))

  # Save plots
  ggsave(file.path(diversity_fig_dir, paste0("shannon_", kingdom, ".png")), plot = shannon_plot, width = 8, height = 6, dpi = 300)
  ggsave(file.path(diversity_fig_dir, paste0("chao1_", kingdom, ".png")), plot = chao1_plot, width = 8, height = 6, dpi = 300)
  ggsave(file.path(diversity_fig_dir, paste0("observed_", kingdom, ".png")), plot = observed_plot, width = 8, height = 6, dpi = 300)

  return(list(shannon_plot, chao1_plot, observed_plot))
}

# Compute and plot diversity for each dataset
diversity_bacteria <- compute_diversity(combined_bacteria, "Bacteria")
diversity_fungi <- compute_diversity(combined_fungi, "Fungi")
diversity_virus <- compute_diversity(combined_virus, "Virus")

# Print plots
print(diversity_bacteria)
print(diversity_fungi)
print(diversity_virus)

```


#CREATE BRAY-CURTIS PCOA PLOTS OF THE BACTERIA, VIRAL, AND FUNGAL SUBSET READS FROM THE PLUSPF DATABASE
```{r}
# Load required libraries
library(ggplot2)
library(vegan)
library(dplyr)  # For data manipulation
library(janitor) # Clean column names
library(ggrepel) # For sample labels

combined_bacteria_transposed_PF <- as.data.frame(t(combined_bacteria))
combined_bacteria_transposed_PF$Sample <- rownames(combined_bacteria_transposed_PF)
rownames(combined_bacteria_transposed_PF) <- NULL
combined_bacteria_transposed_PF <- combined_bacteria_transposed_PF[, c("Sample", setdiff(colnames(combined_bacteria_transposed_PF), "Sample"))]

metadata_IL10$Sample <- rownames(metadata_IL10)  

pcoa_input_bacteria_PF <- merge(combined_bacteria_transposed_PF, metadata_IL10, by = "Sample")
pcoa_numeric_bacteria_PF <- pcoa_input_bacteria_PF[, !(colnames(pcoa_input_bacteria_PF) %in% c("Sample", "Treatment", "Sex"))]
pcoa_numeric_bacteria_PF <- pcoa_numeric_bacteria_PF %>% clean_names()
pcoa_numeric_bacteria_PF <- as.matrix(pcoa_numeric_bacteria_PF)
pcoa_numeric_bacteria_PF <- pcoa_numeric_bacteria_PF / rowSums(pcoa_numeric_bacteria_PF)
bray_curtis_dist_bacteria <- vegdist(pcoa_numeric_bacteria_PF, method="bray")



pcoa_result_bacteria_PF <- cmdscale(bray_curtis_dist_bacteria, k=2, eig=TRUE)  # Compute eigenvalues
pcoa_scores_bacteria_PF <- as.data.frame(pcoa_result_bacteria_PF$points)
colnames(pcoa_scores_bacteria_PF) <- c("PC1", "PC2")
pcoa_scores_bacteria_PF$Sample <- combined_bacteria_transosed_PF$Sample
pcoa_scores_bacteria_PF <- merge(pcoa_scores_bacteria_PF, metadata_IL10, by = "Sample")

# Calculate variance explained
pc1_variance_bacteria_PF <- round(100 * (pcoa_result_bacteria_PF$eig[1] / sum(pcoa_result_bacteria_PF$eig)), 1)
pc2_variance_bacteria_PF <- round(100 * (pcoa_result_bacteria_PF$eig[2] / sum(pcoa_result_bacteria_PF$eig)), 1)

# Print variance explained
cat("PC1 Variance:", pc1_variance_bacteria_PF, "%\n")
cat("PC2 Variance:", pc2_variance_bacteria_PF, "%\n")


# Define colors for Treatment groups
ann_colors <- list(Group = c(
  "ZnD" = "#E69F00",
  "ZnA" = "#56B4E9",
   "ZnS" = "#009E73",
  "Chow" = "#F0E442"
))


# PCoA Plot: Treatment
pcoa_treatment_bacteria_PF <- ggplot(pcoa_scores_bacteria_PF, aes(x = PC1, y = PC2, color = Treatment, fill = Treatment)) +
  geom_point(size = 3, alpha = 0.8) +
  stat_ellipse(geom = "polygon", alpha = 0.2) +  # Add filled ellipses
  scale_color_manual(values = ann_colors$Group) +
  scale_fill_manual(values = ann_colors$Group) +
  #geom_text_repel(aes(label = Sample), size = 3, max.overlaps = 15) +  # Add sample labels
  theme_minimal() +
  labs(
    title = "PCoA Bacteria: Treatment (Bray-Curtis)",
    x = paste0("PC1 (", pc1_variance_bacteria_PF, "%)"),
    y = paste0("PC2 (", pc2_variance_bacteria_PF, "%)"),
    color = "Treatment"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    panel.background = element_rect(fill = "white", color = "black"),
    plot.title = element_text(face = "bold", size = 14),  # Bold title
    axis.title.x = element_text(face = "bold", size = 12),  # Bold x-axis label
    axis.title.y = element_text(face = "bold", size = 12)   # Bold y-axis label
  )
  theme(legend.position = "right", panel.background = element_rect(fill = "white", color = "black"))

# Save and print plot
ggsave("/Users/savitasastry/Downloads/Zinc_Deficiency_SS/Metagenomics/figures/PCA/pcoa_treatment_Bacteria_PF.jpeg", plot = pcoa_treatment_bacteria_PF, width = 8, height = 6, dpi = 300)
pcoa_treatment_bacteria_PF
```

```{r}
# PCoA for Fungi
combined_fungi_transposed_PF <- as.data.frame(t(combined_fungi))
combined_fungi_transposed_PF$Sample <- rownames(combined_fungi_transposed_PF)
rownames(combined_fungi_transposed_PF) <- NULL
combined_fungi_transposed_PF <- combined_fungi_transposed_PF[, c("Sample", setdiff(colnames(combined_fungi_transposed_PF), "Sample"))]
pcoa_input_fungi_PF <- merge(combined_fungi_transposed_PF, metadata_DSS13_c, by = "Sample")

pcoa_numeric_fungi_PF <- pcoa_input_fungi_PF[, !(colnames(pcoa_input_fungi_PF) %in% c("Sample", "Treatment", "Sex", "Cage", "Location"))]
pcoa_numeric_fungi_PF <- pcoa_numeric_fungi_PF %>% clean_names()
pcoa_numeric_fungi_PF <- as.matrix(pcoa_numeric_fungi_PF)
pcoa_numeric_fungi_PF <- pcoa_numeric_fungi_PF / rowSums(pcoa_numeric_fungi_PF)

bray_curtis_dist_fungi <- vegdist(pcoa_numeric_fungi_PF, method="bray")
pcoa_result_fungi_PF <- cmdscale(bray_curtis_dist_fungi, k=2, eig=TRUE)  # Compute eigenvalues

pcoa_scores_fungi_PF <- as.data.frame(pcoa_result_fungi_PF$points)
colnames(pcoa_scores_fungi_PF) <- c("PC1", "PC2")
pcoa_scores_fungi_PF$Sample <- combined_fungi_transposed_PF$Sample
pcoa_scores_fungi_PF <- merge(pcoa_scores_fungi_PF, metadata_DSS13_c, by = "Sample")

# Calculate variance explained
pc1_variance_fungi_PF <- round(100 * (pcoa_result_fungi_PF$eig[1] / sum(pcoa_result_fungi_PF$eig)), 1)
pc2_variance_fungi_PF <- round(100 * (pcoa_result_fungi_PF$eig[2] / sum(pcoa_result_fungi_PF$eig)), 1)

# PCoA Plot for Fungi
pcoa_treatment_fungi_PF <- ggplot(pcoa_scores_fungi_PF, aes(x = PC1, y = PC2, color = Treatment, fill = Treatment)) +
  geom_point(size = 3, alpha = 0.8) +
  stat_ellipse(geom = "polygon", alpha = 0.2) +
  scale_color_manual(values = ann_colors$Group) +
  scale_fill_manual(values = ann_colors$Group) +
  theme_minimal() +
  labs(
    title = "PCoA Fungi: Treatment (Bray-Curtis)",
    x = paste0("PC1 (", pc1_variance_fungi_PF, "%)"),
    y = paste0("PC2 (", pc2_variance_fungi_PF, "%)"),
    color = "Treatment"
  ) +
  theme(
    legend.position = "right",
    plot.title = element_text(face = "bold", size = 14),
    axis.title.x = element_text(face = "bold", size = 12),
    axis.title.y = element_text(face = "bold", size = 12)
  )

# Save and print plot
ggsave("/Users/savitasastry/Downloads/Zinc_Deficiency_SS/Metagenomics_DSS13/1_16_25/figures/PCA/pcoa_treatment_Fungi_PF.jpeg",
       plot = pcoa_treatment_fungi_PF, width = 8, height = 6, dpi = 300)

pcoa_treatment_fungi_PF

```

```{r}
# PCoA for Virus
combined_virus_transposed_PF <- as.data.frame(t(combined_virus))
combined_virus_transposed_PF$Sample <- rownames(combined_virus_transposed_PF)
rownames(combined_virus_transposed_PF) <- NULL
combined_virus_transposed_PF <- combined_virus_transposed_PF[, c("Sample", setdiff(colnames(combined_virus_transposed_PF), "Sample"))]
pcoa_input_virus_PF <- merge(combined_virus_transposed_PF, metadata_DSS13_c, by = "Sample")

pcoa_numeric_virus_PF <- pcoa_input_virus_PF[, !(colnames(pcoa_input_virus_PF) %in% c("Sample", "Treatment", "Sex", "Cage", "Location"))]
pcoa_numeric_virus_PF <- pcoa_numeric_virus_PF %>% clean_names()
pcoa_numeric_virus_PF <- as.matrix(pcoa_numeric_virus_PF)
pcoa_numeric_virus_PF <- pcoa_numeric_virus_PF / rowSums(pcoa_numeric_virus_PF)

bray_curtis_dist_virus <- vegdist(pcoa_numeric_virus_PF, method="bray")
pcoa_result_virus_PF <- cmdscale(bray_curtis_dist_virus, k=2, eig=TRUE)  # Compute eigenvalues

pcoa_scores_virus_PF <- as.data.frame(pcoa_result_virus_PF$points)
colnames(pcoa_scores_virus_PF) <- c("PC1", "PC2")
pcoa_scores_virus_PF$Sample <- combined_virus_transposed_PF$Sample
pcoa_scores_virus_PF <- merge(pcoa_scores_virus_PF, metadata_DSS13_c, by = "Sample")

# Calculate variance explained
pc1_variance_virus_PF <- round(100 * (pcoa_result_virus_PF$eig[1] / sum(pcoa_result_virus_PF$eig)), 1)
pc2_variance_virus_PF <- round(100 * (pcoa_result_virus_PF$eig[2] / sum(pcoa_result_virus_PF$eig)), 1)

# PCoA Plot for Virus
pcoa_treatment_virus_PF <- ggplot(pcoa_scores_virus_PF, aes(x = PC1, y = PC2, color = Treatment, fill = Treatment)) +
  geom_point(size = 3, alpha = 0.8) +
  stat_ellipse(geom = "polygon", alpha = 0.2) +
  scale_color_manual(values = ann_colors$Group) +
  scale_fill_manual(values = ann_colors$Group) +
  theme_minimal() +
  labs(
    title = "PCoA Virus: Treatment (Bray-Curtis)",
    x = paste0("PC1 (", pc1_variance_virus_PF, "%)"),
    y = paste0("PC2 (", pc2_variance_virus_PF, "%)"),
    color = "Treatment"
  ) +
  theme(
    legend.position = "right",
    plot.title = element_text(face = "bold", size = 14),
    axis.title.x = element_text(face = "bold", size = 12),
    axis.title.y = element_text(face = "bold", size = 12)
  )

# Save and print plot
ggsave("/Users/savitasastry/Downloads/Zinc_Deficiency_SS/Metagenomics_DSS13/1_16_25/figures/PCA/pcoa_treatment_Virus_PF.jpeg",
       plot = pcoa_treatment_virus_PF, width = 8, height = 6, dpi = 300)



```

#DIFFERENTIAL EXPRESSION ANALYSIS USING Maaslin2 FOR ALL TAXONOMIC LEVELS, SEPARATED BY BACTERIA, FUNGI, AND VIRUSES

```{r}
library(phyloseq)
library(data.table)
library(readr)
library(tibble)
library(dplyr)

# Load Data
combined_bacteria_S <- read_csv("/Users/savitasastry/Downloads/Zinc_Deficiency_SS/Metagenomics/IL10_Proj/combined_bacteria.csv", col_names = TRUE) %>% 
  column_to_rownames(var = colnames(.)[1]) %>%
  rename_with(~ gsub("_report", "", .))
colnames(combined_bacteria_S) <- gsub("-", ".", colnames(combined_bacteria_S))

combined_virus_S <- read_csv("/Users/savitasastry/Downloads/Zinc_Deficiency_SS/Metagenomics/IL10_Proj/combined_virus.csv", col_names = TRUE) %>% 
  column_to_rownames(var = colnames(.)[1]) %>%
  rename_with(~ gsub("_report", "", .))
colnames(combined_virus_S) <- gsub("-", ".", colnames(combined_virus_S))

combined_fungi_S <- read_csv("/Users/savitasastry/Downloads/Zinc_Deficiency_SS/Metagenomics/IL10_Proj/combined_fungi.csv", col_names = TRUE) %>% 
  column_to_rownames(var = colnames(.)[1]) %>%
  rename_with(~ gsub("_report", "", .))
colnames(combined_fungi_S) <- gsub("-", ".", colnames(combined_fungi_S))

# Load Metadata
metadata_phy_IL10 <- metadata_IL10
rownames(metadata_phy_IL10) <- metadata_phy_IL10$Sample
#metadata_phy_IL10$Treatment <- factor(metadata_phy_IL10$Treatment, levels = c("ZnA", "Chow", "ZnD", "ZnS"))

# 🔹 **Filter Metadata for Chow vs. ZnS**
metadata_chow_vs_zns <- metadata_phy_IL10 %>% filter(Treatment %in% c("Chow", "ZnS"))
metadata_chow_vs_zns$Treatment <- factor(metadata_chow_vs_zns$Treatment, levels = c("Chow", "ZnS"))
metadata_zna_vs_znd <- metadata_phy_IL10 %>% filter(Treatment %in% c("ZnA", "ZnD"))
metadata_zna_vs_znd$Treatment <- factor(metadata_zna_vs_znd$Treatment, levels = c("ZnA", "ZnD"))

# 🔹 **Filter Count Tables for Chow vs. ZnS**
bacteria_chow_vs_zns <- combined_bacteria_S[, colnames(combined_bacteria_S) %in% metadata_chow_vs_zns$Sample]
fungi_chow_vs_zns <- combined_fungi_S[, colnames(combined_fungi_S) %in% metadata_chow_vs_zns$Sample]
virus_chow_vs_zns <- combined_virus_S[, colnames(combined_virus_S) %in% metadata_chow_vs_zns$Sample]

# 🔹 **Filter Count Tables for ZnA vs. ZnD**
bacteria_zna_vs_znd <- combined_bacteria_S[, colnames(combined_bacteria_S) %in% metadata_zna_vs_znd$Sample]
fungi_zna_vs_znd <- combined_fungi_S[, colnames(combined_fungi_S) %in% metadata_zna_vs_znd$Sample]
virus_zna_vs_znd <- combined_virus_S[, colnames(combined_virus_S) %in% metadata_zna_vs_znd$Sample]

# Convert Metadata to `phyloseq` format
#metadata_chow_vs_zns <- sample_data(metadata_chow_vs_zns)
#metadata_zna_vs_znd <- sample_data(metadata_zna_vs_znd)

# **Create Phyloseq Objects**
physeq_bacteria_chow_vs_zns <- phyloseq(otu_table(bacteria_chow_vs_zns, taxa_are_rows = TRUE), metadata_chow_vs_zns)
physeq_fungi_chow_vs_zns <- phyloseq(otu_table(fungi_chow_vs_zns, taxa_are_rows = TRUE), metadata_chow_vs_zns)
physeq_virus_chow_vs_zns <- phyloseq(otu_table(virus_chow_vs_zns, taxa_are_rows = TRUE), metadata_chow_vs_zns)

physeq_bacteria_zna_vs_znd <- phyloseq(otu_table(bacteria_zna_vs_znd, taxa_are_rows = TRUE), metadata_zna_vs_znd)
physeq_fungi_zna_vs_znd <- phyloseq(otu_table(fungi_zna_vs_znd, taxa_are_rows = TRUE), metadata_zna_vs_znd)
physeq_virus_zna_vs_znd <- phyloseq(otu_table(virus_zna_vs_znd, taxa_are_rows = TRUE), metadata_zna_vs_znd)

```

```{r}
library(phyloseq)
library(data.table)
library(readr)
library(tibble)

combined_bacteria_S <- read_csv("/Users/savitasastry/Downloads/Zinc_Deficiency_SS/Metagenomics/IL10_Proj/combined_bacteria.csv", col_names = TRUE) %>% 
  column_to_rownames(var = colnames(.)[1]) %>%
  rename_with(~ gsub("_report", "", .))
colnames(combined_bacteria_S) <- gsub("-", ".", colnames(combined_bacteria_S))

combined_virus_S <- read_csv("/Users/savitasastry/Downloads/Zinc_Deficiency_SS/Metagenomics/IL10_Proj/combined_virus.csv", col_names = TRUE) %>% 
  column_to_rownames(var = colnames(.)[1]) %>%
  rename_with(~ gsub("_report", "", .))
colnames(combined_virus_S) <- gsub("-", ".", colnames(combined_virus_S))

combined_fungi_S <- read_csv("/Users/savitasastry/Downloads/Zinc_Deficiency_SS/Metagenomics/IL10_Proj/combined_fungi.csv", col_names = TRUE) %>% 
  column_to_rownames(var = colnames(.)[1]) %>%
  rename_with(~ gsub("_report", "", .))
colnames(combined_fungi_S) <- gsub("-", ".", colnames(combined_fungi_S))

combined_bacteria_S
metadata_IL10
combined_virus_S
combined_fungi_S

metadata_phy_IL10 <- metadata_IL10
rownames(metadata_phy_IL10) <- metadata_IL10$Sample
metadata_phy_IL10$Treatment <- factor(metadata_phy_IL10$Treatment, levels = c("ZnA", "Chow", "ZnD", "ZnS"))

physeq_ref_control_bacteria <- phyloseq(
  otu_table(combined_bacteria_S, taxa_are_rows = TRUE),  # Taxa abundance matrix
  sample_data(metadata_phy_IL10)                        # Metadata
)
physeq_ref_control_fungi <- phyloseq(
  otu_table(combined_fungi_S, taxa_are_rows = TRUE),  # Taxa abundance matrix
  sample_data(metadata_phy_IL10)                        # Metadata
)
physeq_ref_control_virus <- phyloseq(
  otu_table(combined_virus_S, taxa_are_rows = TRUE),  # Taxa abundance matrix
  sample_data(metadata_phy_IL10)                        # Metadata
)

```


#DIFFERENTIAL EXPRESSION FOR BACTERIA FUNGI AND VIRUSES (NOT SEPARATED BY LOCATION)

```{r}
library(Maaslin2)

# **Transform Count Data**
bacteria_chow_vs_zns_transposed <- t(bacteria_chow_vs_zns)
bacteria_chow_vs_zns_transposed <- as.matrix(bacteria_chow_vs_zns_transposed)
mode(bacteria_chow_vs_zns_transposed) <- "numeric"
bacteria_chow_vs_zns_transposed <- as.data.frame(bacteria_chow_vs_zns_transposed)
#bacteria_chow_vs_zns_transposed <- as.data.frame(bacteria_chow_vs_zns_transposed)



bacteria_zna_vs_znd_transposed <- as.data.frame(t(bacteria_zna_vs_znd))

fungi_chow_vs_zns_transposed <- as.data.frame(t(fungi_chow_vs_zns))
fungi_zna_vs_znd_transposed <- as.data.frame(t(fungi_zna_vs_znd))

virus_chow_vs_zns_transposed <- as.data.frame(t(virus_chow_vs_zns))
virus_zna_vs_znd_transposed <- as.data.frame(t(virus_zna_vs_znd))
metadata_chow_vs_zns <- as.data.frame(metadata_chow_vs_zns)

# **Run Maaslin2 for Chow vs. ZnS**
masalin_results_bacteria_chow_vs_zns <- Maaslin2(
  bacteria_chow_vs_zns_transposed,
  metadata_chow_vs_zns,
  output = "/Users/savitasastry/Downloads/Zinc_Deficiency_SS/Metagenomics/IL10_Proj/DifferentialExpression/masalin_results_bacteria_chow_vs_zns", 
  min_prevalence = 0.05,
  fixed_effects = c("Treatment")
)

masalin_results_fungi_chow_vs_zns <- Maaslin2(
  fungi_chow_vs_zns_transposed,
  metadata_chow_vs_zns,
  output = "/Users/savitasastry/Downloads/Zinc_Deficiency_SS/Metagenomics/IL10_Proj/DifferentialExpression/masalin_results_fungi_chow_vs_zns", 
  fixed_effects = c("Treatment")
)

masalin_results_virus_chow_vs_zns <- Maaslin2(
  virus_chow_vs_zns_transposed,
  metadata_chow_vs_zns,
  output = "/Users/savitasastry/Downloads/Zinc_Deficiency_SS/Metagenomics/IL10_Proj/DifferentialExpression/masalin_results_virus_chow_vs_zns", 
  fixed_effects = c("Treatment")
)

# **Run Maaslin2 for ZnA vs. ZnD**
masalin_results_bacteria_zna_vs_znd <- Maaslin2(
  bacteria_zna_vs_znd_transposed,
  metadata_zna_vs_znd,
  output = "/Users/savitasastry/Downloads/Zinc_Deficiency_SS/Metagenomics/IL10_Proj/DifferentialExpression/masalin_results_bacteria_zna_vs_znd", 
  min_prevalence = 0.05,
  fixed_effects = c("Treatment")
)

masalin_results_fungi_zna_vs_znd <- Maaslin2(
  fungi_zna_vs_znd_transposed,
  metadata_zna_vs_znd,
  output = "/Users/savitasastry/Downloads/Zinc_Deficiency_SS/Metagenomics/IL10_Proj/DifferentialExpression/masalin_results_fungi_zna_vs_znd", 
  fixed_effects = c("Treatment")
)

masalin_results_virus_zna_vs_znd <- Maaslin2(
  virus_zna_vs_znd_transposed,
  metadata_zna_vs_znd,
  output = "/Users/savitasastry/Downloads/Zinc_Deficiency_SS/Metagenomics/IL10_Proj/DifferentialExpression/masalin_results_virus_zna_vs_znd", 
  fixed_effects = c("Treatment")
)

```
```{r}
str(bacteria_chow_vs_zns_transposed)
str(metadata_chow_vs_zns)
str(bacteria_transposed)
str(metadata_phy_IL10)
```







```{r}
library(Maaslin2)

combined_virus_S
as.data.frame(t(combined_virus_S))

virus_transposed
metadata_phy_ZnD5

#INPUT FOR VIRUSES
virus_transposed <- t(combined_virus_S)
#colnames(virus_transposed) <- paste0("T", seq_len(ncol(virus_transposed)))
virus_transposed <- as.matrix(virus_transposed)
mode(virus_transposed) <- "numeric"
virus_transposed <- as.data.frame(virus_transposed)

#INPUT FOR BACTERIA
bacteria_transposed <- t(combined_bacteria_S)
#colnames(virus_transposed) <- paste0("T", seq_len(ncol(virus_transposed)))
#bacteria_transposed <- as.matrix(bacteria_transposed)
#mode(bacteria_transposed) <- "numeric"
bacteria_transposed <- as.data.frame(bacteria_transposed)

#INPUT FOR FUNGI 
fungi_transposed <- t(combined_fungi_S)
fungi_transposed <- as.matrix(fungi_transposed)
#colnames(fungi_transposed) <- paste0("T", seq_len(ncol(fungi_transposed)))
mode(fungi_transposed) <- "numeric"
fungi_transposed <- as.data.frame(fungi_transposed)

# 3️⃣ Run Maaslin2 with the fixed dataset
masalin_results_bacteria_S <- Maaslin2(
  bacteria_transposed,
  metadata_phy_IL10,
  output = "/Users/savitasastry/Downloads/Zinc_Deficiency_SS/Metagenomics/IL10_Proj/DifferentialExpression/masalin_results_bacteria_S", 
  min_prevalence = 0.05,
  fixed_effects = c("Treatment")
)

masalin_results_fungi_S <- Maaslin2(
  fungi_transposed,
  metadata_phy_IL10,
  output = "/Users/savitasastry/Downloads/Zinc_Deficiency_SS/Metagenomics/IL10_Proj/DifferentialExpression/masalin_results_fungi_S", 
  fixed_effects = c("Treatment")
)

masalin_results_virus_S <- Maaslin2(
  virus_transposed,
  metadata_phy_IL10,
  output = "/Users/savitasastry/Downloads/Zinc_Deficiency_SS/Metagenomics/IL10_Proj/DifferentialExpression/masalin_results_virus_S", 
  fixed_effects = c("Treatment")
)

metadata_phy_IL10
metadata_IL10
```

#DIFFERENTIAL EXPRESSION SEPARATED BY COLON, INTESTNE, AND CECUM AND BACTERIA/FUNGI/VIRUSES
```{r}
# Load Maaslin2
library(Maaslin2)
library(dplyr)

# 1️⃣ **Filter metadata for each location**
metadata_colon <- metadata_phy_ZnD5 %>% filter(Location == "colon")
metadata_intestine <- metadata_phy_ZnD5 %>% filter(Location == "intestine")
metadata_cecum <- metadata_phy_ZnD5 %>% filter(Location == "cecum")

# 2️⃣ **Subset expression data to match metadata for each location**
filter_data_by_location <- function(data, metadata) {
  data_filtered <- data[rownames(data) %in% metadata$Sample, ]
  return(data_filtered)
}

# **Filter Bacteria**
bacteria_colon <- filter_data_by_location(bacteria_transposed, metadata_colon)
bacteria_intestine <- filter_data_by_location(bacteria_transposed, metadata_intestine)
bacteria_cecum <- filter_data_by_location(bacteria_transposed, metadata_cecum)

# **Filter Fungi**
fungi_colon <- filter_data_by_location(fungi_transposed, metadata_colon)
fungi_intestine <- filter_data_by_location(fungi_transposed, metadata_intestine)
fungi_cecum <- filter_data_by_location(fungi_transposed, metadata_cecum)

# **Filter Virus**
virus_colon <- filter_data_by_location(virus_transposed, metadata_colon)
virus_intestine <- filter_data_by_location(virus_transposed, metadata_intestine)
virus_cecum <- filter_data_by_location(virus_transposed, metadata_cecum)

# 3️⃣ **Run Maaslin2 for Each Intestinal Site**
run_maaslin2 <- function(data, metadata, output_dir, label) {
  Maaslin2(
    input_data = data,
    input_metadata = metadata,
    output = paste0(output_dir, "_", label),
    min_prevalence = 0.05,
    fixed_effects = c("Treatment")
  )
}

# **Define output directory**
output_dir <- "/Users/savitasastry/Downloads/Zinc_Deficiency_SS/Metagenomics/DifferentialExpression"

# **Run Maaslin2 for Bacteria**
run_maaslin2(bacteria_colon, metadata_colon, output_dir, "bacteria_colon")
run_maaslin2(bacteria_intestine, metadata_intestine, output_dir, "bacteria_intestine")
run_maaslin2(bacteria_cecum, metadata_cecum, output_dir, "bacteria_cecum")

# **Run Maaslin2 for Fungi**
run_maaslin2(fungi_colon, metadata_colon, output_dir, "fungi_colon")
run_maaslin2(fungi_intestine, metadata_intestine, output_dir, "fungi_intestine")
run_maaslin2(fungi_cecum, metadata_cecum, output_dir, "fungi_cecum")

# **Run Maaslin2 for Virus**
run_maaslin2(virus_colon, metadata_colon, output_dir, "virus_colon")
run_maaslin2(virus_intestine, metadata_intestine, output_dir, "virus_intestine")
run_maaslin2(virus_cecum, metadata_cecum, output_dir, "virus_cecum")


```

#GENERATE HEATMAPS FOR BACTERIA, FUNGI, AND VIRUSES, SEPARATED BY LOCATION SELECTING THE TOP 30 MOST SIGNIFICANT DIFFERENTIALLY EXPRESSED TAXA FROM MAASLIN2

```{r}
# Load required libraries
library(ggplot2)
library(ComplexHeatmap)
library(dplyr)
library(tidyr)
library(RColorBrewer)

# Define input paths
base_path <- "/Users/savitasastry/Downloads/Zinc_Deficiency_SS/Metagenomics/IL10_Proj/DifferentialExpression/"

# Define output path for heatmaps
heatmap_output_dir <- "/Users/savitasastry/Downloads/Zinc_Deficiency_SS/Metagenomics/IL10_Proj/figures/Heatmaps"
if (!dir.exists(heatmap_output_dir)) dir.create(heatmap_output_dir, recursive = TRUE)

# Load metadata
metadata <- metadata_IL10_phy

# Function to filter data by location
filter_data_by_location <- function(data, location) {
  # Identify samples belonging to the specified location
  location_samples <- metadata$Sample[metadata$Location == location]
  # Subset data based on location-specific samples
  data_subset <- data[, colnames(data) %in% location_samples]
  return(data_subset)
}

# Function to generate heatmaps
generate_heatmap <- function(category, data, location) {
  
  # Define Maaslin2 results path
  maaslin_results_path <- file.path(base_path, paste0("DifferentialExpression_", category, "_", location, "/all_results.tsv"))
  
  # Check if Maaslin2 results exist
  if (!file.exists(maaslin_results_path)) {
    cat("Skipping:", maaslin_results_path, "- File not found\n")
    return(NULL)
  }
  
  # Load Maaslin2 results
  maaslin_results <- read.delim(maaslin_results_path)
  
  # **Fix species names in Maaslin2 results** (replace '.' with ' ')
  maaslin_results$feature <- gsub("\\.", " ", maaslin_results$feature)

  # Select the **top 30 significant species** sorted by p-value
  top_species <- maaslin_results %>%
    arrange(pval) %>%
    head(30) %>%
    pull(feature)

  # Subset data based on the **top 30 significant species**
  data_subset <- filter_data_by_location(data, location)
  data_subset <- data_subset[rownames(data_subset) %in% top_species, ]

  # Check if we have species left after filtering
  if (nrow(data_subset) == 0) {
    cat("Warning: No matching top species found in the dataset for Heatmap of Top 30 Significant", category, "-", location, "\n")
    return(NULL)
  }

  # Normalize relative abundance per species
  data_subset <- t(apply(data_subset, 1, function(x) x / sum(x)))

  # Define color palette
  heatmap_colors <- colorRampPalette(rev(brewer.pal(9, "RdBu")))(100)

  # Generate heatmap
  heatmap_plot <- Heatmap(
    data_subset,
    name = "Relative Abundance",
    col = heatmap_colors,
    row_names_side = "left",
    cluster_rows = TRUE,
    cluster_columns = TRUE,
    show_row_names = TRUE,
    show_column_names = TRUE,
    column_names_gp = gpar(fontsize = 10),
    row_names_gp = gpar(fontsize = 10),
    heatmap_legend_param = list(title = "Relative Abundance"),
    column_title = paste("Heatmap of Top 30 Significant", category, "-", location)
  )

  # Save the heatmap
  output_file <- file.path(heatmap_output_dir, paste0("heatmap_", category, "_", location, ".png"))
  png(output_file, width = 1200, height = 1200, res = 150)
  draw(heatmap_plot)
  dev.off()
  
  cat("Generated heatmap for:", category, location, "\n")
}

# Run heatmap generation for **each category and location**
categories <- c("bacteria", "fungi", "virus")
locations <- c("colon", "intestine", "cecum")

for (category in categories) {
  for (location in locations) {
    if (category == "bacteria") {
      generate_heatmap(category, combined_bacteria_S, location)
    } else if (category == "fungi") {
      generate_heatmap(category, combined_fungi_S, location)
    } else {
      generate_heatmap(category, combined_virus_S, location)
    }
  }
}

cat("All heatmaps saved in:", heatmap_output_dir, "\n")

```

```{r}
# Load required libraries
library(ggplot2)
library(ComplexHeatmap)
library(dplyr)
library(tidyr)
library(RColorBrewer)

# Define input paths
base_path <- "/Users/savitasastry/Downloads/Zinc_Deficiency_SS/Metagenomics/DifferentialExpression"

# Define output path for heatmaps
heatmap_output_dir <- "/Users/savitasastry/Downloads/Zinc_Deficiency_SS/Metagenomics/figures/Heatmaps"
if (!dir.exists(heatmap_output_dir)) dir.create(heatmap_output_dir, recursive = TRUE)

# Load metadata
metadata <- metadata_phy_ZnD5

# Function to filter data by location
filter_data_by_location <- function(data, location) {
  # Identify samples belonging to the specified location
  location_samples <- metadata$Sample[metadata$Location == location]
  # Subset data based on location-specific samples
  data_subset <- data[, colnames(data) %in% location_samples]
  return(data_subset)
}

# Function to generate heatmaps with ZnD and ZnA grouping
generate_heatmap <- function(category, data, location) {
  
  # Define Maaslin2 results path
  maaslin_results_path <- file.path(base_path, paste0("DifferentialExpression_", category, "_", location, "/all_results.tsv"))
  
  # Check if Maaslin2 results exist
  if (!file.exists(maaslin_results_path)) {
    cat("Skipping:", maaslin_results_path, "- File not found\n")
    return(NULL)
  }
  
  # Load Maaslin2 results
  maaslin_results <- read.delim(maaslin_results_path)
  
  # **Fix species names in Maaslin2 results** (replace '.' with ' ')
  maaslin_results$feature <- gsub("\\.", " ", maaslin_results$feature)

  # Select the **top 30 significant species** sorted by p-value
  top_species <- maaslin_results %>%
    arrange(pval) %>%
    head(30) %>%
    pull(feature)

  # Subset data based on the **top 30 significant species**
  data_subset <- filter_data_by_location(data, location)
  data_subset <- data_subset[rownames(data_subset) %in% top_species, ]

  # Check if we have species left after filtering
  if (nrow(data_subset) == 0) {
    cat("Warning: No matching top species found in the dataset for Heatmap of Top 30 Significant", category, "-", location, "\n")
    return(NULL)
  }

  # Normalize relative abundance per species
  data_subset <- t(apply(data_subset, 1, function(x) x / sum(x)))

  # **Order columns by Treatment (ZnD | ZnA)**
  treatment_order <- metadata %>% 
    filter(Sample %in% colnames(data_subset)) %>%
    arrange(Treatment)  # Ensure ZnD is first, then ZnA

  data_subset <- data_subset[, treatment_order$Sample]

  # Define **annotation for Treatment groups**
  treatment_colors <- c("ZnD" = "#E69F00", "ZnA" = "#56B4E9")
  column_annotation <- HeatmapAnnotation(
    Treatment = treatment_order$Treatment,
    col = list(Treatment = treatment_colors)
  )

  # Define color palette for heatmap
  heatmap_colors <- colorRampPalette(rev(brewer.pal(9, "RdBu")))(100)

  # Generate heatmap
  heatmap_plot <- Heatmap(
    data_subset,
    name = "Relative Abundance",
    col = heatmap_colors,
    row_names_side = "left",
    cluster_rows = TRUE,
    cluster_columns = FALSE,  # Keep fixed order ZnD -> ZnA
    show_row_names = TRUE,
    show_column_names = TRUE,
    column_names_gp = gpar(fontsize = 10),
    row_names_gp = gpar(fontsize = 10),
    heatmap_legend_param = list(title = "Relative Abundance"),
    column_title = paste("Heatmap of Top 30 Significant", category, "-", location),
    top_annotation = column_annotation  # Add Treatment annotation
  )

  # Save the heatmap
  output_file <- file.path(heatmap_output_dir, paste0("heatmap_", category, "_", location, ".png"))
  png(output_file, width = 1200, height = 1200, res = 150)
  draw(heatmap_plot)
  dev.off()
  
  cat("Generated heatmap for:", category, location, "\n")
}

# Run heatmap generation for **each category and location**
categories <- c("bacteria", "fungi", "virus")
locations <- c("colon", "intestine", "cecum")

for (category in categories) {
  for (location in locations) {
    if (category == "bacteria") {
      generate_heatmap(category, combined_bacteria_S, location)
    } else if (category == "fungi") {
      generate_heatmap(category, combined_fungi_S, location)
    } else {
      generate_heatmap(category, combined_virus_S, location)
    }
  }
}

cat("All heatmaps saved in:", heatmap_output_dir, "\n")

```


```{r}
unique(metadata_phy_ZnD5$Location)
metadata_phy_ZnD5$Location <- tolower(metadata_phy_ZnD5$Location)
table(metadata_phy_ZnD5$Location)
setdiff(metadata_phy_ZnD5$Sample, colnames(combined_bacteria_S))

```
```{r}
# Load Maaslin2 results for bacteria (adjust for fungi/viruses)
maaslin_results_path <- "/Users/savitasastry/Downloads/Zinc_Deficiency_SS/Metagenomics/DifferentialExpression/DifferentialExpression_bacteria_colon/all_results.tsv"
maaslin_results <- read.delim(maaslin_results_path)

# Get the top 30 species sorted by p-value
top_species <- maaslin_results %>%
  arrange(pval) %>%
  head(30) %>%
  pull(feature)

top_species
combined_bacteria_S

# Check how many of these are in the dataset
missing_species <- setdiff(top_species, rownames(combined_bacteria_S))

cat("Number of missing species in dataset:", length(missing_species), "\n")
cat("Missing species:\n", paste(missing_species, collapse = ", "), "\n")

```

